---
title: "razzo pipeline"
author: "Giovanni Laudanno and Richel J.C. Bilderbeek"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{razzo pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  \usepackage[T1]{fontenc}
  \usepackage{lmodern}
---

## Legend

Abbreviation|Full
---|---
nLTT|normalized lineages through time
BD|Birth Death Model
MBD|Multiple Birth Death Model

This is the pipeline to follow for the Razzo project. 

Our aim is to compare two nLTT distributions related to two tree posteriors: one descending from MBD alignments and another one from BD alignments. Both posteriors are generated using a BD prior in BEAST.

## Overview

Update our dependencies:

```{r message=FALSE, install_dependencies}
if (!require(ape)) {install.packages("ape")}
if (!require(mbd)) {devtools::install_github("Giappo/mbd", quiet = TRUE)}
devtools::install_github("ropensci/beautier", quiet = TRUE)
devtools::install_github("ropensci/tracerer", quiet = TRUE)
devtools::install_github("ropensci/beastier", quiet = TRUE)
devtools::install_github("ropensci/mauricer", quiet = TRUE)
devtools::install_github("ropensci/babette", quiet = TRUE)
devtools::install_github("richelbilderbeek/mcbette", quiet = TRUE)
devtools::install_github("richelbilderbeek/pirouette", quiet = TRUE)
```

Load the libraries needed:

```{r load_libraries}
library(mbd)
library(razzo)
library(ggplot2)
```

We will work in this folder:

```{r create_working_folder}
super_folder_name <- tempdir()
project_folder_name <- file.path(super_folder_name, "razzo_project") 
dir.create(path = project_folder_name, recursive = TRUE)
```

## Step 1: Create parameter files

Here we create all parameter files:

```{r create_parameter_files}
parameters_filenames <- create_parameters_files(
  project_folder_name = project_folder_name
)
```

## Step 2: Run the parameter files

```
for (i in seq_along(parameters_filenames)) {
  parameters_filename <- parameters_filenames[i]
  razzo::run_razzo(parameters_filename)
}
```

## Step 3: figure 1

```
create_fig_1(project_folder_name = project_folder_name)
```

## Step 4: what is the effect of MBD on the error?

The nLTT statistics of the BD tree shows the baseline error:
the Bayesian analysis assumes BD, and that tree *is* BD.

```
mbd_nltts <- utils::read.csv(mbd_nltt_filenames[1])$x
bd_nltts <- utils::read.csv(bd_nltt_filenames[1])$x

df <- data.frame(
  model = c(rep("MBD", length(mbd_nltts)), rep("BD", length(bd_nltts))),
  nltt = c(mbd_nltts, bd_nltts),
  stringsAsFactors = TRUE  
)

plot <- ggplot(
  data = df,
  aes(x = nltt, fill = model)
) + ggplot2::scale_x_continuous(limits = c(0.0, 1.0))

plot + geom_histogram(binwidth = 0.01, alpha = 0.5)
plot + geom_density(alpha = 0.5)
```


## Step 5: measure Effective Sample Sizes

We already had a sneak-peek at the ESSes.
In this step, we measure the ESSes of all files and
combine it into one file:

```
esses_filename <- create_esses_file(project_folder_name = project_folder_name)
knitr::kable(utils::read.csv(esses_filename)[-1])
```

## Step 6: compare marginal likelihoods

We already had a sneak-peek at the ESSes.
In this step, we measure the ESSes of all files and
combine it into one file:

```
if (rappdirs::app_dir()$os != "win") {
  marg_liks_file <- create_marg_liks_file(
    project_folder_name = project_folder_name
  )
  knitr::kable(utils::read.csv(marg_liks_file)[-1])
}
```

# Appendix

### First MBD tree:

```
ape::plot.phylo(ape::read.tree(file = mbd_tree_filenames[2]))
```


### First BD tree:

```
ape::plot.phylo(ape::read.tree(file = bd_tree_filenames[2]))
```

### First MBD alignment

```
image(ape::read.FASTA(file = mbd_alignment_filenames[1]))
```

### First BD alignment

```
image(ape::read.FASTA(file = bd_alignment_filenames[1]))
```

### First MBD posterior's trees

```
babette::plot_densitree(tracerer::parse_beast_trees(mbd_alignment_filenames[[1]][1]))
```

### First MBD posterior's ESSes

```
knitr::kable(
  tracerer::calc_esses(
    tracerer::parse_beast_log(mbd_alignment_filenames[[1]][2]), 
    sample_interval = open_parameters_file(
      parameters_filename[1]
    )$sample_interval
  )
)
```


### First BD posterior's trees

```
babette::plot_densitree(tracerer::parse_beast_trees(bd_alignment_filenames[[1]][1]))
```

### First BD posterior's ESSes

Plot the Effective Sample Sizes (must be at least 200 in manuscript):

```
knitr::kable(
  tracerer::calc_esses(
    tracerer::parse_beast_log(bd_alignment_filenames[[1]][2]), 
    sample_interval = open_parameters_file(
      parameters_filename[1]
    )$sample_interval
  )
)
```

### First MBD errors

```
ggplot(
  data = data.frame(nltt = utils::read.csv(mbd_nltt_filenames[1])$x),
  aes(x = nltt)
) + geom_histogram(binwidth = 0.01) + 
  ggplot2::scale_x_continuous(limits = c(0.0, 1.0))
```


### First BD errors

```
ggplot(
  data = data.frame(nltt = utils::read.csv(bd_nltt_filenames[1])$x),
  aes(x = nltt)
) + geom_histogram(binwidth = 0.01) + 
  ggplot2::scale_x_continuous(limits = c(0.0, 1.0))
```

### First MBD marginal likelihoods

```
if (rappdirs::app_dir()$os != "win") {
  bd_marg_lik_filenames <- rep(NA, length(parameters_filenames))
  for (i in seq_along(parameters_filenames)) {
    bd_marg_lik_filenames[i] <- create_bd_marg_lik_file(
      parameters_filename = parameters_filenames[i]
    )
  }
  knitr::kable(utils::read.csv(bd_marg_lik_filenames[1])[-1])
}
```

### First BD marginal likelihoods

```
if (rappdirs::app_dir()$os != "win") {
  bd_marg_lik_filenames <- rep(NA, length(parameters_filenames))
  for (i in seq_along(parameters_filenames)) {
    bd_marg_lik_filenames[i] <- create_bd_marg_lik_file(
      parameters_filename = parameters_filenames[i]
    )
  }
  knitr::kable(utils::read.csv(bd_marg_lik_filenames[1])[-1])
}
```

