---
title: "razzo pipeline"
author: "Giovanni Laudanno and Richel J.C. Bilderbeek"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{razzo pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  \usepackage[T1]{fontenc}
  \usepackage{lmodern}
---

## Legend

Abbreviation|Full
---|---
nLTT|normalized lineages through time
BD|Birth Death Model
MBD|Multiple Birth Death Model

This is the pipeline to follow for the Razzo project. 

Our aim is to compare two nLTT distributions related to two tree posteriors: one descending from MBD alignments and another one from BD alignments. Both posteriors are generated using a BD prior in BEAST.

## Overview

Update our dependencies:

```{r message=FALSE, install_dependencies}
if (1 == 2) {
  if (!require(ape)) {install.packages("ape")}
  if (!require(mbd)) {devtools::install_github("Giappo/mbd", quiet = TRUE)}
  devtools::install_github("ropensci/mauricer", quiet = TRUE)
  devtools::install_github("ropensci/babette", quiet = TRUE)
  devtools::install_github("richelbilderbeek/mcbette", quiet = TRUE)
  devtools::install_github("richelbilderbeek/pirouette", quiet = TRUE)
}
```

Load the libraries needed:

```{r load_libraries}
library(razzo)
```

We will work in this folder:

```{r create_working_folder}
super_folder_name <- rappdirs::user_cache_dir()
project_folder_name <- file.path(super_folder_name, "razzo_project") 
dir.create(path = project_folder_name, recursive = TRUE, showWarnings = FALSE)
```

## Step 1: Create parameter files

Here we create all testing parameter files, which are only a few:

```{r create_parameter_files}
parameters_filenames <- create_files_razzo_paramses(
  project_folder_name = project_folder_name,
  experiment_type = "test"
)
```

## Step 2: Run the parameter files

Then we run razzo for each parameter setting

```{r run_files}
for (parameters_filename in parameters_filenames) {
  razzo::run_razzo_from_file(parameters_filename)
}
```

## Step 3: figure 1

```{r}
if (rappdirs::app_dir()$os != "win") {
  plots <- create_fig_1(project_folder_name)
  for (plot in plots) {
    print(plot)
  }
}
```

## Step 4: what is the effect of MBD on the error?

```
# create_fig_2(project_folder_name)
```


The nLTT statistics of the BD tree shows the baseline error:
the Bayesian analysis assumes BD, and that tree *is* BD.

```
mbd_nltts <- utils::read.csv(mbd_nltt_filenames[1])$x
bd_nltts <- utils::read.csv(bd_nltt_filenames[1])$x

df <- data.frame(
  model = c(rep("MBD", length(mbd_nltts)), rep("BD", length(bd_nltts))),
  nltt = c(mbd_nltts, bd_nltts),
  stringsAsFactors = TRUE  
)

plot <- ggplot(
  data = df,
  aes(x = nltt, fill = model)
) + ggplot2::scale_x_continuous(limits = c(0.0, 1.0))

plot + geom_histogram(binwidth = 0.01, alpha = 0.5)
plot + geom_density(alpha = 0.5)
```


## Step 5: measure Effective Sample Sizes

We already had a sneak-peek at the ESSes.
In this step, we measure the ESSes of all files and
combine it into one file:

```
esses_filename <- create_esses_file(project_folder_name = project_folder_name)
knitr::kable(utils::read.csv(esses_filename)[-1])
```

## Step 6: compare marginal likelihoods

We already had a sneak-peek at the ESSes.
In this step, we measure the ESSes of all files and
combine it into one file:

```
if (rappdirs::app_dir()$os != "win") {
  marg_liks_file <- create_marg_liks_file(
    project_folder_name = project_folder_name
  )
  knitr::kable(utils::read.csv(marg_liks_file)[-1])
}
```

# Appendix

Here we show the results of the first experiment only:

```{r}
parameters_filename <- parameters_filenames[1]
razzo_params <- readRDS(parameters_filename)
```
### First MBD tree:

```{r}
ape::plot.phylo(ape::read.tree(file = razzo_params$misc_params$tree_filename))
```


### First twin tree:

```{r}

ape::plot.phylo(
  ape::read.tree(
    file = razzo_params$pir_params$twinning_params$twin_tree_filename
  )
)
```

### First MBD alignment

```{r}
image(
  ape::read.FASTA(
    file = razzo_params$pir_params$alignment_params$fasta_filename
  )
)
```

### First twin alignment

```{r}
image(
  ape::read.FASTA(
    file = razzo_params$pir_params$twinning_params$twin_alignment_filename
  )
)
```

### First MBD posterior's trees of generative model

```{r}
babette::plot_densitree(
  tracerer::parse_beast_trees(
    razzo_params$pir_params$experiments[[1]]$beast2_options$output_trees_filenames
  )
)
```

### First MBD posterior's trees of best candidate model

```{r}
# babette::plot_densitree(
#   tracerer::parse_beast_trees(
#     razzo_params$pir_params$experiments[[2]]$beast2_options$output_trees_filenames
#   )
# )
```

### First MBD posterior's ESSes

```{r}
knitr::kable(
  tracerer::calc_esses(
    tracerer::parse_beast_log(razzo_params$pir_params$experiments[[1]]$beast2_options$output_log_filename), 
    sample_interval = razzo_params$pir_params$experiments[[1]]$inference_model$mcmc$store_every
  )
)
```


### First twin posterior's trees of generative model

```{r}
babette::plot_densitree(
  tracerer::parse_beast_trees(
    pirouette::to_twin_filename(
      razzo_params$pir_params$experiments[[1]]$beast2_options$output_trees_filenames
    )
  )
)
```

### First twin posterior's trees of best candidate model

```{r}
# babette::plot_densitree(
#   tracerer::parse_beast_trees(
#     pirouette::to_twin_filename(
#       razzo_params$pir_params$experiments[[2]]$beast2_options$output_trees_filenames
#     )
#   )
# )
```
### First BD posterior's ESSes

Plot the Effective Sample Sizes (must be at least 200 in manuscript):

```{r}
# knitr::kable(
#   tracerer::calc_esses(
#     tracerer::parse_beast_log(
#       pirouette::to_twin_filename(
#         razzo_params$pir_params$experiments[[2]]$beast2_options$output_log_filename)
#     ), 
#     sample_interval = razzo_params$pir_params$experiments[[2]]$inference_model$mcmc$store_every
#   )
# )
```

### First MBD errors

```
ggplot(
  data = data.frame(nltt = utils::read.csv(mbd_nltt_filenames[1])$x),
  aes(x = nltt)
) + geom_histogram(binwidth = 0.01) + 
  ggplot2::scale_x_continuous(limits = c(0.0, 1.0))
```


### First BD errors

```
ggplot(
  data = data.frame(nltt = utils::read.csv(bd_nltt_filenames[1])$x),
  aes(x = nltt)
) + geom_histogram(binwidth = 0.01) + 
  ggplot2::scale_x_continuous(limits = c(0.0, 1.0))
```

### First MBD marginal likelihoods

```
if (rappdirs::app_dir()$os != "win") {
  bd_marg_lik_filenames <- rep(NA, length(parameters_filenames))
  for (i in seq_along(parameters_filenames)) {
    bd_marg_lik_filenames[i] <- create_bd_marg_lik_file(
      parameters_filename = parameters_filenames[i]
    )
  }
  knitr::kable(utils::read.csv(bd_marg_lik_filenames[1])[-1])
}
```

### First BD marginal likelihoods

```
if (rappdirs::app_dir()$os != "win") {
  bd_marg_lik_filenames <- rep(NA, length(parameters_filenames))
  for (i in seq_along(parameters_filenames)) {
    bd_marg_lik_filenames[i] <- create_bd_marg_lik_file(
      parameters_filename = parameters_filenames[i]
    )
  }
  knitr::kable(utils::read.csv(bd_marg_lik_filenames[1])[-1])
}
```

